Scoring Patients

### 2018-02-20 Patient_Report.Rmd
Updating the pipeline to make it more generic

```{r setup, include=FALSE}
library(ggplot2)
library(ggbio)
library(scales)
library(GGally)
library(plyr)
library(RColorBrewer)
library(dplyr)
library(pander)
library(ggdendro)
library(reshape2)
library(GenomicRanges)
library(gridExtra)
library(tibble)

@@

@@

if (length(list.files(plot.dir)) <= 0)
  stop(paste("No analysis files found in", plot.dir ))

@@

#head(patient.info)

@@

@@

#pander(arrange(patient.data, Status), justify='left', caption='Patient information')
```

# Description of '`r dataset`' data set

```{r cors, echo=F, message=T, warning=F, fig.height=6, fig.width=6, fig.align='left', include=F}
## Cluster the samples by coverage then pull out regions

## Thresholds
gamma2 = 250
min.probes=67 # equiv to 1MB
sd.threshold = 0.08  # cov by numbers of probes
gain.threshold = 1.1; loss.threshold = 0.9

cachefile = paste(data, '/Analysis/', dataset, '_patients.Rdata', sep='')
if ( file.exists(cachefile) ) {
  load(cachefile, verbose=T)
} else {
  patient.data = list(); lowprobes = c()

  for (patient.name in sum.patient.data$Hospital.Research.ID  )  {
    message(patient.name)
    patient.plot.dir = paste(plot.dir, patient.name, sep='/')

    if (!file.exists(paste(patient.plot.dir, "/",patient.name,"_segmentedCoverage_fitted_gamma",gamma2,".txt",sep=""))) {
      warning(paste("Missing segmented file for", patient.name))
      next
    }

    segvals = read.table(paste(patient.plot.dir, "/",patient.name,"_segmentedCoverage_fitted_gamma",gamma2,".txt",sep=""),sep="\t",stringsAsFactors=F,header=T)
    head(segvals)

    missing = c( setdiff(colnames(segvals)[-(1:5)], subset(patient.info, Hospital.Research.ID == patient.name)$Samplename),
                 setdiff(subset(patient.info, Hospital.Research.ID == patient.name)$Samplename, colnames(segvals)[-(1:5)]) )

    mr = with(removed.samples, which(Hospital.Research.ID == patient.name & Samplename %in% missing))
    if (length(mr) > 0) {
      missing = missing[-which(missing %in% removed.samples[mr,]$Samplename)]
      segvals = segvals[,-which(colnames(segvals) %in% removed.samples[mr,]$Samplename) ]
    }

    if (length(missing) > 0) {
      warning(paste("Missing", length(missing), "sample(s) of", length(subset(patient.info, Patient == patient.name)$Samplename), "in patient", patient.name))
      #next
    }

    probes = length(which(segvals$n.probes < min.probes))
    message( paste(probes, ' probes (', round(probes/nrow(segvals),2)*100, '%)',' below the minimum probe count (',min.probes,')', sep=''))
    if (probes/nrow(segvals) >= 0.75) lowprobes = c(lowprobes, patient.name)

    segvals = segvals[segvals$n.probes >= min.probes,]
    dim(segvals)

    # ORDER segvals by endoscopy year etc
    cur.pt =  arrange( subset(patient.info, Hospital.Research.ID == patient.name), Endoscopy.Year, Pathology)
    segvals = segvals[, c(colnames(segvals[1:5]), intersect(colnames(segvals), cur.pt$Samplename))]

    if (nrow(segvals) < 100) {
      warning(paste(patient.name, 'has fewer than 100 genomic regions with segmented values:', nrow(segvals)))
      #next
    }

    patient.data[[patient.name]][['seg.vals']] = segvals
    patient.data[[patient.name]][['info']] = cur.pt

    sample.cols = grep('^D\\d+_D\\d+', colnames(segvals))

    if (length(sample.cols) > 0) {
      if (length(sample.cols) > 2) {  
        HC = hclust(dist(t(segvals[,sample.cols])))
        gg1 = ggdendrogram(HC, rotate=T) + labs(title = paste(patient.name,": fitted segment coverage"),x="")
        #print(gg1)
      }

      # Normalize  (value-mean(value))/sd(value)
      normalised.segvals = segvals[,sample.cols]
      if (length(sample.cols) > 1) {  
        for(c in 1:nrow(normalised.segvals)) 
          normalised.segvals[c,] = (normalised.segvals[c,]-mean(unlist(normalised.segvals[c,])))/sd(unlist(normalised.segvals[c,]))
        patient.data[[patient.name]][['norm.seg.vals']] = normalised.segvals

        HC = hclust(dist(t(normalised.segvals)))
        gg2 = ggdendrogram(HC, rotate=T) + labs(title=paste(patient.name,": normalised fitted segment coverage", x=""))
        #print(gg2)

        no.samples = ncol(normalised.segvals)
        no.rows = ceiling(no.samples/3)

        gg3 = ggplot( melt(normalised.segvals), aes(x=value) ) + 
          facet_wrap(~variable,scales = "free_x") + 
          geom_histogram( fill=I("blue"),col=I("lightblue"),binwidth=0.3) + labs(title=paste(patient.name,'normalised segment coverage'), x='', y='')
      }

      # Variable Regions can only be called relative to each sample so if there's only one sample we can't make a call
      if (length(sample.cols) > 1) {
        sds = apply(segvals[,sample.cols],1,sd)
        means = apply(segvals[,sample.cols],1,mean)
        CofV = sds/means

        variable.region.indices = which(sds >= sd.threshold & segvals$n.probes >= min.probes)
        variable.regions = segvals[variable.region.indices, intersect(patient.data[[patient.name]]$info$Samplename, colnames(segvals))] 
        patient.data[[patient.name]][['variable.regions']] = cbind(segvals[variable.region.indices,c(1:5)],variable.regions )
      }

      patient = arrange(subset(patient.info, Hospital.Research.ID == patient.name),  Endoscopy.Year, Pathology, Samplename)
    }
  }
save(patient.data, file=cachefile)
}
print(paste("Missing patients:", paste(setdiff(sum.patient.data$Hospital.Research.ID, names(patient.data)), collapse=', ')))

sum.patient.data = subset(sum.patient.data, Hospital.Research.ID %in% names(patient.data))
table(sum.patient.data$Status)

## Check that all of the samples are there
patient.info = arrange(patient.info, Status, Patient, Hospital.Research.ID, Endoscopy.Year,Samplename)
msgs = sapply(names(patient.data), function(pt) {
  missing = which(!patient.data[[pt]]$info$Samplename %in% colnames(patient.data[[pt]]$seg.vals))
  if (length(missing) > 0) {
    msg = paste(pt,": Missing samples - ", paste(patient.data[[pt]]$info$Samplename[missing], collapse=', ') , sep='')
    msg = paste(msg, "\nFrom batch: ", paste(unique(patient.data[[pt]]$info$Batch.Name[missing]), collapse=', '))
    warning(msg)
  }
})
for (n in msgs) {
  if (!is.null(n))
    print(n)
}

example.pt = which(with(sum.patient.data, Status == 'P' & total.samples > 2))[5]
key =  sum.patient.data[example.pt, 'Hospital.Research.ID' ]
example.pt.id = sum.patient.data[example.pt, 'Patient']
## Note the number of probes per patient/sample
```
\pagebreak

@@

1. Segmented values (progressor): `r example.pt` 
```{r}
head(patient.data[[ key ]]$seg.vals)
```

```{r echo=F, warning=F, message=F, fig.height=10, fig.width=10}
ggplot( melt(patient.data[[key]]$seg.vals[,-(1:5)]), aes(x=value) ) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram( fill=I("blue"),col=I("lightblue"), binwidth=0.03) + 
  labs(title=paste('Patient', example.pt.id,' segmented value distribution'), x='', y='')
```

2. Row (e.g. per-region across timepoints) normalized values from the matrix above
```{r}
head(patient.data[[key]]$norm.seg.vals)
```

```{r echo=F, warning=F, message=F, fig.height=10, fig.width=10}
ggplot( melt(patient.data[[key]]$norm.seg.vals), aes(x=value) ) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram( fill=I("blue"),col=I("lightblue"), binwidth=0.3) + labs(title=paste('Patient', example.pt.id,'normalized segmented value distribution'), x='', y='')

@@

```{r general, warning=F, echo=F, message=F}
gen = do.call(rbind, lapply(patient.data, function(df) {
  cbind('n.segments'=nrow(df[['seg.vals']]), 
        'median.seg.len'=median(df$seg.vals[,c('end.pos')] - df$seg.vals[,c('start.pos')]),
        'mean.seg.len'=mean(df$seg.vals[,c('end.pos')] - df$seg.vals[,c('start.pos')]), 
        'sd.seg.len'=sd(df$seg.vals[,c('end.pos')] - df$seg.vals[,c('start.pos')]))
}))
rownames(gen)  = names(patient.data)

@@

ggplot(sum.patient.data) + geom_boxplot( aes(y=median.seg.len, x=Status, group=Status, fill=Status) ) +
  labs(title="Median segment length by patient status")
```

@@

var.seg.lengths = lapply(patient.data, function(df) {
  cbind('chr'=df$variable.regions$chrom, 'start'=df$variable.regions$start.pos, 'end'=df$variable.regions$end.pos, 'seg.len'=df$variable.regions$end.pos - df$variable.regions$start.pos)
})

segment.lengths = lapply(patient.data, function(df) {
  cbind('chr'=df$seg.vals$chrom, 'seg.len'=df$seg.vals$end.pos - df$seg.vals$start.pos)
})

sum.var = rbind(
  'Progressors'=do.call(rbind.data.frame, var.seg.lengths[prog.pt]) %>% summarise( 'n.segments' = length(seg.len), 'median.length'=median(seg.len), 'mean.length'=mean(seg.len), 'sd'=sd(seg.len)),
  'Non-Progressors'=do.call(rbind.data.frame, var.seg.lengths[np.pt]) %>% summarise( 'n.segments' = length(seg.len), 'median.length'=median(seg.len), 'mean.length'=mean(seg.len), 'sd'=sd(seg.len)))

pander(sum.var, caption='Variable regions between progressors and non-progressors')

sum.seg = rbind(
  'Progressors'=do.call(rbind.data.frame, segment.lengths[prog.pt]) %>% summarise( 'n.segments' = length(seg.len), 'median.length'=median(seg.len), 'mean.length'=mean(seg.len), 'sd'=sd(seg.len)),
  'Non-Progressors'=do.call(rbind.data.frame, segment.lengths[np.pt]) %>% summarise( 'n.segments' = length(seg.len), 'median.length'=median(seg.len), 'mean.length'=mean(seg.len), 'sd'=sd(seg.len)))

pander(sum.var, caption='Segment stats between progressors and non-progressors')
```

df = rbind(cbind(do.call(rbind.data.frame, var.seg.lengths[prog.pt]), 'status'='Progressor'), cbind(do.call(rbind.data.frame, var.seg.lengths[np.pt]), 'status'='Non-Progressor'))
lenWT = wilcox.test(subset(df, status == 'Progressor')$seg.len, subset(df, status == 'Non-Progressor')$seg.len)

ggplot(df, aes(status, seg.len)) + geom_jitter(aes(color=status), alpha=0.5, show.legend = F, width=.2) +
  geom_boxplot(aes(group=status, fill=status), show.legend=F, outlier.colour = NA) + 
  geom_label(data=subset(df, status == 'Progressor'), aes(status, max(seg.len)-2*sd(seg.len), label=paste('median\n',scientific(median(seg.len),3),sep='') )) +
  geom_label(data=subset(df, status == 'Non-Progressor'), aes(status, max(seg.len), label=paste('median\n',scientific(median(seg.len),3),sep=''))) +
  labs(y='Variable segment length', x='', title='Variable regions', caption=paste('Wilcox p-value', signif(lenWT$p.value, digits=3) ))

ggplot(df, aes(status, seg.len)) + 
  geom_jitter(aes(color=status), show.legend=F) + geom_violin(aes(group=status, fill=status), show.legend=F) +
  geom_label(data=subset(df, status == 'Progressor'), aes(status, max(seg.len)-2*sd(seg.len),label=paste('median\n',scientific( median(seg.len) , 3),sep='') ))  +
    geom_label(data=subset(df, status == 'Non-Progressor'), aes(status, max(seg.len), label=paste('median\n',scientific( median(seg.len) , 3),sep=''))) +
    labs(y='Variable segment length', x='', title='Variable regions', caption=paste('Wilcox p-value', signif(lenWT$p.value, digits=3) ))

@@

```{r, echo=F, warning=F, message=F}
df = rbind(cbind.data.frame('n'=sapply(var.seg.lengths[prog.pt], nrow), 'status'='Progressor'), cbind.data.frame('n'=sapply(var.seg.lengths[np.pt], nrow), 'status'='Non-Progressor'))
nWT = wilcox.test(subset(df, status == 'Progressor')$n, subset(df, status == 'Non-Progressor')$n)

ggplot(df, aes(status, n, group=status)) + geom_jitter(aes(color=status), width=0.2) +
  geom_boxplot(aes(fill=status), position = 'dodge', show.legend = F, outlier.colour = NA) +
  geom_label(data=subset(df, status == 'Progressor'), 
             aes(status, max(n)-2*sd(n), label=paste('median\n',sprintf("%.1f",median(n)),sep=''), fill=status ), color='white' ) +
  geom_label(data=subset(df, status == 'Non-Progressor'), 
             aes(status, max(n)+2*sd(n), label=paste('median\n',sprintf("%.1f",median(n)),sep=''), fill=status), color='white') +
  labs(y='Number of segments', title="Total number of segments", caption=paste('Wilcox p-value', signif(nWT$p.value, digits=3)) ) +
  theme(legend.position = 'none')

ggplot(df, aes(status, n, group=status)) + geom_jitter(aes(color=status), width=0.2) +
  geom_violin(aes(fill=status, group=status)) +
  geom_label(data=subset(df, status == 'Progressor'), 
             aes(status, max(n)-2*sd(n), label=paste('median\n',sprintf("%.1f",median(n)),sep=''), fill=status ), color='white' ) +
  geom_label(data=subset(df, status == 'Non-Progressor'), 
             aes(status, max(n)+2*sd(n), label=paste('median\n',sprintf("%.1f",median(n)),sep=''), fill=status), color='white') +
  labs(y='Number of segments', title="Total number of segments", caption=paste('Wilcox p-value', signif(nWT$p.value, digits=3)) ) +
  theme(legend.position = 'none')

@@

```{r echo=F, message=F, warning=F, fig.height=8, fig.width=20}
n.var.regions = sapply(var.seg.lengths, nrow)

@@

sum.patient.data$Num.Var.Regions = as.vector(sapply(var.seg.lengths[sum.patient.data$Patient], nrow))

@@

var.seg.lengths = lapply(var.seg.lengths, function(x) merge(x, chr.info, by='chr') )

# All segments
segment.lengths = lapply(segment.lengths, function(x) merge(x, chr.info, by='chr') )

# Genome coverage
genome.coverage = sapply(segment.lengths, function(x) {
  sum(as.numeric(x[['seg.len']]))/sum(as.numeric(chr.info$length))
})

# No difference in coverage
covtt = wilcox.test(genome.coverage[prog.pt], genome.coverage[np.pt])

var.seg.lengths = lapply(var.seg.lengths, function(x) cbind(x, 'chr.ratio'=x[['seg.len']]/x[['length']]))

# Segment lengths, variable regions
vsl = do.call(rbind.data.frame, lapply(patient.data, function(df) {
  if (is.null(df$variable.regions) || nrow(df$variable.regions) <= 0) return(NULL)

  mc = grep('^D', colnames(df$variable.regions))

@@

    cbind.data.frame( 'chr'=df$variable.regions$chrom, 'start'=df$variable.regions$start.pos, 'end'=df$variable.regions$end.pos, meanCN, 'status'= df$info$Status[1])
  }))
vsl = vsl[-which(vsl$meanCN == 0),]

@@

ggplot(vslm, aes(x=length)) + 
  facet_grid(status~chr) +
  geom_rect(data=vslm, aes(xmin=start, xmax=end, ymin=0, ymax=value, fill=value)) +
  scale_fill_gradient(low = "green3", high = "red3", name='') + 
  labs(title="Variable region gains & losses", x='Chromosomes', y='Mean adjusted segmentation value') +
  theme(axis.text.x=element_blank(),axis.ticks.x=element_blank(), legend.position='bottom')

@@

```{r echo=F, message=F, warning=F}
# Variable region length by ratio of chromosome
chr.ratios = lapply(var.seg.lengths, function(x) x[['seg.len']]/x[['length']])

melted = melt(chr.ratios)
melted = merge(sum.patient.data[,c('Hospital.Research.ID', 'Status')], melted, by.x='Hospital.Research.ID', by.y='L1')

chr.ratio.tt = t.test(unlist(chr.ratios[prog.pt]), unlist(chr.ratios[np.pt]))

ggplot(melted, aes(Status, value, fill=Status)) + geom_jitter(aes(color=Status), width=0.2) + geom_boxplot(outlier.colour = NA) + 
  geom_label(data=subset(melted, Status == 'P'), aes(x=Status, y=max(value)-sd(value), label=scientific(median(value),3) ), color='white', show.legend = F ) +
  geom_label(data=subset(melted, Status == 'NP'), aes(x=Status, y=max(value)-sd(value), label=scientific(median(value),3) ), color='white' , show.legend = F) +
  labs(title="Variable region length",subtitle=paste("p-value", signif(chr.ratio.tt$p.value, digits=3)), y='Segment Length/Chr length')

ggplot(melted, aes(Status, value, fill=Status)) + geom_jitter(aes(color=Status), width=0.2) + geom_violin() + 
  geom_label(data=subset(melted, Status == 'P'), aes(x=Status, y=max(value)-sd(value), label=scientific(median(value),3) ), color='white' , show.legend = F) +
  geom_label(data=subset(melted, Status == 'NP'), aes(x=Status, y=max(value)-sd(value), label=scientific(median(value),3) ), color='white' , show.legend = F) +
  labs(title="Variable region length",subtitle=paste("p-value", signif(chr.ratio.tt$p.value, digits=3)), y='Segment Length/Chr length')

var.seg.lengths =  lapply(var.seg.lengths, function(df) {
   ddply(df, .(chr, length), summarise, seg.lengths = sum(seg.len), chr.ratio = sum(seg.len)/length[1])
})

# Genome coverage of the variable regions
pcov = median(sapply(var.seg.lengths[prog.pt], function(x) 
  sum(as.numeric(x[['seg.lengths']]))/sum(as.numeric(x[['length']])) ), na.rm=T)

npcov = median(sapply(var.seg.lengths[np.pt], function(x) 
  sum(as.numeric(x[['seg.lengths']]))/sum(as.numeric(x[['length']])) ), na.rm=T)

@@

long.chr.tt = t.test(unlist(lapply(var.seg.lengths[prog.pt], get.chrs, chrs=c(1:12))), unlist(lapply(var.seg.lengths[np.pt], get.chrs, chrs=c(1:12))))
short.chr.tt = t.test(unlist(lapply(var.seg.lengths[prog.pt], get.chrs, chrs=c(13:22))), unlist(lapply(var.seg.lengths[np.pt], get.chrs, chrs=c(13:22))))

@@

#ftL = fisher.test(df)
```

So it appears that adjusting for chromosome length (looking at the ratio of the chromosome covered by segments rather than the raw bp) a significant difference in the coverage of the genome by CN segments that are variable between progressors and non: t-test p-value `r signif(chr.ratio.tt$p.value, 3)`.  Overall bp coverage of the sWGS is `r round(mean(genome.coverage), 2)*100`%  (P `r round(mean(genome.coverage[prog.pt]),2)`, NP `r round(mean(genome.coverage[np.pt]), 2)`).

Variable regions cover `r round(pcov, 2)*100`% of the the progressors genome, and `r round(npcov, 2)*100`% of non-progressor genomes.

@@

ft = fisher.test((df %>% group_by(status) %>% summarise( 'gains'=sum(gain), 'losses'=sum(loss)))[,c('gains','losses')])

@@

>>>>

```{r setup, include=FALSE}
library(ggplot2)
library(ggbio)
library(scales)
library(GGally)
library(plyr)
library(RColorBrewer)
library(pander)
library(ggdendro)
library(reshape2)
library(GenomicRanges)
library(gridExtra)
#library(tibble)
library(dplyr)

@@

get.var.regions <- function(segvals, valueCols, sd=0.08) {
  if (is.null(valueCols)) {
    warning('No sample columns indicated')
    return(NULL)
  }
#  if (length(sample.cols) > 1) {
  sds = apply(segvals[,valueCols],1,sd)
  means = apply(segvals[,valueCols],1,mean)
  CofV = sds/means

  variable.region.indices = which(sds >= sd)
  segvals[variable.region.indices,] 
#}

}

@@

cleaned = list.files(paste(data, 'Cleaned', sep='/'), pattern='tiled', full.names=T, recursive=T)
rawseg = list.files(paste(data, 'Cleaned', sep='/'), pattern='raw', full.names=T, recursive=T)
normseg = list.files(paste(data, 'Cleaned', sep='/'), pattern='norm', full.names=T, recursive=T)

@@

@@

patient.data = list()
for(hr in sum.patient.data$Hospital.Research.ID) {
  info = subset(patient.info, Hospital.Research.ID == hr)
  info = arrange(info, Endoscopy.Year, Pathology)

  segvals = read.table( grep(hr, rawseg, value=T), sep='\t', header=T)
  if (nrow(info) > 1)
    segvals = cbind(segvals[,c(1:5)], segvals[,intersect(info$Samplename, colnames(segvals))])

  patient.data[[hr]][['info']] = info
  patient.data[[hr]][['seg.vals']] = segvals
  patient.data[[hr]][['norm.seg.vals']] = read.table( grep(hr, normseg, value=T), sep='\t', header=T)

  if (nrow(info) > 1) # Variable regions are called relative to each other so cannot do it on a single sample
    patient.data[[hr]][['variable.regions']] = get.var.regions(segvals, c(6:ncol(segvals)))  
}

@@

#pander(arrange(patient.data, Status), justify='left', caption='Patient information')

example.pt = subset(sum.patient.data, Status == 'P')$Hospital.Research.ID[12]
example.pt.id = subset(sum.patient.data, Status == 'P')$PatientID[12]
```

```{r echo=F, eval=F}

tiled = do.call(rbind, lapply(cleaned, function(f) {
  data.table::fread(f, sep='\t', header=T)
}))

@@

1. Segmented values (progressor): `r example.pt` 
```{r}

head(patient.data[[ example.pt ]]$seg.vals)
```

```{r echo=F, warning=F, message=F, fig.height=10, fig.width=10}
ggplot( melt(patient.data[[example.pt]]$seg.vals[,-(1:5)]), aes(x=value) ) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram( fill=I("blue"),col=I("lightblue"), binwidth=0.03) + 
  labs(title=paste('Patient', example.pt.id,' segmented value distribution'), x='', y='')
```

2. Row (e.g. per-region across timepoints) normalized values from the matrix above
```{r}
head(patient.data[[example.pt]]$norm.seg.vals)
```

```{r echo=F, warning=F, message=F, fig.height=10, fig.width=10}
ggplot( melt(patient.data[[example.pt]]$norm.seg.vals), aes(x=value) ) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram( fill=I("blue"),col=I("lightblue"), binwidth=0.3) + labs(title=paste('Patient', example.pt.id,'normalized segmented value distribution'), x='', y='')

@@

```{r general, warning=F, echo=F, message=F}
hgdSamp = subset(patient.info, Pathology %in% c('HGD','IMC'))$Samplename

gen = do.call(rbind, lapply(patient.data, function(df) {
  cbind('n.segments'=nrow(df[['seg.vals']]), 
        'median.seg.len'=median(df$seg.vals[,c('end.pos')] - df$seg.vals[,c('start.pos')]),
        'mean.seg.len'=mean(df$seg.vals[,c('end.pos')] - df$seg.vals[,c('start.pos')]), 
        'sd.seg.len'=sd(df$seg.vals[,c('end.pos')] - df$seg.vals[,c('start.pos')]))
}))
rownames(gen)  = names(patient.data)

@@

ggplot(sum.patient.data) + geom_boxplot( aes(y=median.seg.len, x=Status, group=Status, fill=Status), show.legend = F ) +
  labs(title="Median segment length by patient status")

@@

var.seg.lengths = do.call(rbind, lapply(patient.data, function(df) { 
  if (is.null(df$variable.regions) || nrow(df$variable.regions) <= 0) return(NULL)
  vr = cbind(df$variable.regions[,c(1:5)], 'Status' = unique(df$info$Status), 'Hospital.Research.ID'=unique(df$info$Hospital.Research.ID))
  vr %>% rowwise() %>% mutate(seg.len = end.pos-start.pos)
}))

segment.lengths = do.call(rbind, lapply(patient.data, function(df) { 
  vr = cbind(df$seg.vals[,c(1:5)], 'Status' = unique(df$info$Status))
  vr %>% rowwise() %>% mutate(seg.len = end.pos-start.pos)
}))

sum.var = ddply(var.seg.lengths, .(Status), summarise, 'n.segments' = length(seg.len), 'median.length'=median(seg.len), 'mean.length'=mean(seg.len), 'sd'=sd(seg.len), 'max'=max(seg.len))
pander(sum.var, caption='Variable regions between progressors and non-progressors')

sum.seg = ddply(segment.lengths, .(Status), summarise, 'n.segments' = length(seg.len), 'median.length'=median(seg.len), 'mean.length'=mean(seg.len), 'sd'=sd(seg.len), 'max'=max(seg.len))
pander(sum.var, caption='Segment stats between progressors and non-progressors')
```

## Excluding HGD

```{r, echo-F}
var.seg.lengthsnohgd = do.call(rbind, lapply(patient.data, function(df) { 
  samples = subset(df$info, Pathology %in% c('BE','ID','LGD'))$Samplename
  cols = which(colnames(df$seg.vals) %in% samples)

  if (length(cols) > 1)
    vr = get.var.regions(df$seg.vals, cols)

  if (!exists('vr') || nrow(vr) <= 0) return(NULL)
  vr = cbind(vr[,c(1:5)], 'Status' = unique(df$info$Status), 'Hospital.Research.ID'=unique(df$info$Hospital.Research.ID))
  vr %>% rowwise() %>% mutate(seg.len = end.pos-start.pos)
}))

sum.varnohgd = ddply(var.seg.lengthsnohgd, .(Status), summarise, 'n.segments' = length(seg.len), 'median.length'=median(seg.len), 'mean.length'=mean(seg.len), 'sd'=sd(seg.len), 'max'=max(seg.len))

pander(sum.varnohgd, caption='Variable regions between progressors and non-progressors NO HGD')

@@

M-W for variable segment length in progressors with/without HGD
```{r}
wilcox.test(subset(var.seg.lengths, Status == 'P')$seg.len, subset(var.seg.lengthsnohgd, Status == 'P')$seg.len)
```

And non-progressors
```{r}
wilcox.test(subset(var.seg.lengths, Status == 'NP')$seg.len, subset(var.seg.lengthsnohgd, Status == 'NP')$seg.len)
```

```{r echo=F, message=F, warning=F}
lenWT = wilcox.test(subset(var.seg.lengths, Status == 'P')$seg.len, subset(var.seg.lengths, Status == 'NP')$seg.len)
ggplot() + geom_jitter(data=var.seg.lengths, aes(Status, seg.len, color=Status), alpha=0.5, show.legend = F, width=.2) +
  geom_boxplot(data=var.seg.lengths, aes(Status, seg.len, group=Status, fill=Status), show.legend=F, outlier.colour = NA) + 
  geom_label(data=sum.var, aes(Status, max, label=paste('median\n',scientific(median.length,3),sep='') )) +
  labs(y='Variable segment length', x='', title='Variable regions', subtitle=paste('Wilcox p-value', signif(lenWT$p.value, digits=3) ))

lenWT = wilcox.test(subset(var.seg.lengthsnohgd, Status == 'P')$seg.len, subset(var.seg.lengthsnohgd, Status == 'NP')$seg.len)
ggplot() + geom_jitter(data=var.seg.lengthsnohgd, aes(Status, seg.len, color=Status), alpha=0.5, show.legend = F, width=.2) +
  geom_boxplot(data=var.seg.lengths, aes(Status, seg.len, group=Status, fill=Status), show.legend=F, outlier.colour = NA) + 
  geom_label(data=sum.varnohgd, aes(Status, max, label=paste('median\n',scientific(median.length,3),sep='') )) +
  labs(y='Variable segment length', x='', title='Variable regions, NO HGD', subtitle=paste('Wilcox p-value', signif(lenWT$p.value, digits=3) ))

vs = rbind(cbind(subset(var.seg.lengths, Status == 'P'), 'HGD'=T),
      cbind(subset(var.seg.lengthsnohgd, Status == 'P'), 'HGD'=F))
sum.vs = ddply(vs, .(HGD), summarise, max = max(seg.len), median.length=median(seg.len), sd=sd(seg.len), seg.len=0 )

lenWT = wilcox.test(subset(vs, HGD)$seg.len, subset(vs, !HGD)$seg.len)
ggplot(data=vs, aes(HGD, seg.len)) + 
  geom_jitter(aes(color=HGD), alpha=0.5, show.legend = F, width=.2) +
  geom_boxplot(aes(group=HGD, fill=HGD), show.legend=F, outlier.colour = NA) + 
  geom_label(data=sum.vs, aes(HGD, max, label=paste('median\n',scientific(median.length,3),sep='') )) +
  labs(y='Variable segment length', x='', title='Variable regions, Excluding HGD/IMC', subtitle=paste('Wilcox p-value', signif(lenWT$p.value, digits=3) ))
```

@@

```{r echo=F, message=F, warning=F, fig.height=8, fig.width=20}
n.var.regions = nrow(var.seg.lengths)

@@

#sum.patient.data$Num.Var.Regions = as.vector(sapply(var.seg.lengths[sum.patient.data$Patient], nrow))

@@

var.seg.lengths = merge(var.seg.lengths, chr.info, by.x='chrom',by.y='chr') 
var.seg.lengthsnohgd = merge(var.seg.lengthsnohgd, chr.info, by.x='chrom',by.y='chr')

# All segments
segment.lengths = merge(segment.lengths, chr.info, by.x='chrom', by.y='chr')

# Genome coverage
segment.lengths = segment.lengths %>% rowwise() %>% mutate(coverage=seg.len/length)

# No difference in coverage
covtt = wilcox.test(subset(segment.lengths, Status == 'P')$coverage,subset(segment.lengths, Status == 'NP')$coverage)

var.seg.lengths = var.seg.lengths %>% rowwise() %>% mutate(coverage=seg.len/length)
var.seg.lengthsnohgd = var.seg.lengthsnohgd %>% rowwise() %>% mutate(coverage=seg.len/length)

getCN <- function(df, path=c('BE','ID','LGD','HGD','IMC')) {
  if (!is.null(df$variable.regions) && nrow(df$variable.regions) > 0) {
    samples = subset(df$info, Pathology %in% path)$Samplename
    mc = intersect(colnames(df$variable.regions), samples)

@@

   return(cbind.data.frame( 'chr'=df$variable.regions$chrom, 'start'=df$variable.regions$start.pos, 'end'=df$variable.regions$end.pos, meanCN, 'status'= df$info$Status[1]))
  } 
return(NULL)
}
# Segment lengths, variable regions
vsl = do.call(rbind.data.frame, lapply(patient.data, getCN, path=c('BE','ID','LGD')))
vsl = vsl[-which(vsl$meanCN == 0),]

@@

ggplot(vslm, aes(x=length)) + 
  facet_grid(status~chr) +
  geom_rect(data=vslm, aes(xmin=start, xmax=end, ymin=0, ymax=value, fill=value)) +
  scale_fill_gradient(low = "green3", high = "red3", name='') + 
  labs(title="Variable region gains & losses", subtitle='Excluding HGD/IMC', x='Chromosomes', y='Mean adjusted segmentation value') +
  theme(axis.text.x=element_blank(),axis.ticks.x=element_blank(), legend.position='bottom')

vsl = do.call(rbind.data.frame, lapply(patient.data, getCN))
vsl = vsl[-which(vsl$meanCN == 0),]

ggplot(vslm, aes(x=length)) + 
  facet_grid(status~chr) +
  geom_rect(data=vslm, aes(xmin=start, xmax=end, ymin=0, ymax=value, fill=value)) +
  scale_fill_gradient(low = "green3", high = "red3", name='') + 
  labs(title="Variable region gains & losses", subtitle='All samples', x='Chromosomes', y='Mean adjusted segmentation value') +
  theme(axis.text.x=element_blank(),axis.ticks.x=element_blank(), legend.position='bottom')

@@

```{r echo=F, message=F, warning=F}
# Variable region length by ratio of chromosome
melted = melt(var.seg.lengths, measure.vars=c('coverage'))
chr.ratio.tt = t.test(subset(var.seg.lengths, Status == 'P')$coverage, subset(var.seg.lengths, Status == 'NP')$coverage)

sm = ddply(melted, .(Status), summarise, median.cov=median(value), max=max(value), value=0)

ggplot(melted, aes(Status, value)) + geom_jitter(aes(color=Status), width=0.2) + geom_violin(alpha=0.3) + 
  geom_label(data=sm, aes(Status, max, label=paste('Median coverage\n', scientific(median.cov, 3), sep=''))) +
  labs(title="Variable region coverage, All samples",subtitle=paste("p-value", signif(chr.ratio.tt$p.value, digits=3)), y='Variable region Segment Length/Chr length')

melted = melt(var.seg.lengthsnohgd, measure.vars=c('coverage'))
chr.ratio.tt = t.test(subset(var.seg.lengthsnohgd, Status == 'P')$coverage, subset(var.seg.lengthsnohgd, Status == 'NP')$coverage)

sm = ddply(melted, .(Status), summarise, median.cov=median(value), max=max(value), value=0)
ggplot(melted, aes(Status, value)) + geom_jitter(aes(color=Status), width=0.2) + geom_violin(alpha=0.3) + 
  geom_label(data=sm, aes(Status, max, label=paste('Median coverage\n', scientific(median.cov, 3), sep=''))) +
  labs(title="Variable region coverage, Excluding HGD/IMC",subtitle=paste("p-value", signif(chr.ratio.tt$p.value, digits=3)), y='Variable region Segment Length/Chr length')

# Genome coverage of the variable regions

cov = ddply(var.seg.lengths, .(Status), summarise, coverage=sum(as.numeric(seg.len))/sum(as.numeric(length)))

# pcov = median(sapply(var.seg.lengths[prog.pt], function(x) 
#   sum(as.numeric(x[['seg.lengths']]))/sum(as.numeric(x[['length']])) ), na.rm=T)
# 
# npcov = median(sapply(var.seg.lengths[np.pt], function(x) 
#   sum(as.numeric(x[['seg.lengths']]))/sum(as.numeric(x[['length']])) ), na.rm=T)

@@

long.chr.tt = t.test(subset(var.seg.lengths, Status == 'P' & chrom %in% c(1:12))$coverage,subset(var.seg.lengths, Status == 'NP' & chrom %in% c(1:12))$coverage)
short.chr.tt = t.test(subset(var.seg.lengths, Status == 'P' & chrom %in% c(13:22))$coverage,subset(var.seg.lengths, Status == 'NP' & chrom %in% c(13:22))$coverage)

@@

#ftL = fisher.test(df)

#So it appears that adjusting for chromosome length (looking at the ratio of the chromosome covered by segments rather than the raw bp) a significant difference in the coverage of the genome by CN segments that are variable between progressors and non: t-test p-value `r signif(chr.ratio.tt$p.value, 3)`.  Overall bp coverage of the sWGS is `r round(mean(genome.coverage), 2)*100`%  (P `r round(mean(genome.coverage[prog.pt]),2)`, NP `r round(mean(genome.coverage[np.pt]), 2)`).

#Variable regions cover `r round(pcov, 2)*100`% of the the progressors genome, and `r round(npcov, 2)*100`% of non-progressor genomes.
```

@@

ft = fisher.test(ddply(df, .(status), summarise,'gains'=sum(gain), 'losses'=sum(loss)) [,c('gains','losses')])

@@

But only in the variable regions EXCLUDING HGD/IMC
